\documentclass[11pt]{article}
\usepackage[a4paper,margin=1in]{geometry}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{longtable}

\title{Mental Math --- System Architecture (v1)}
\date{}

\begin{document}

\maketitle

\section{Purpose \& Scope}

This document defines the current architecture of the Mental Math application as implemented in the repository.
The stack is intentionally fixed for v1 and optimized for:

\begin{itemize}[leftmargin=*]
	\item Fast product iteration with one TypeScript codebase.
	\item Server-rendered authenticated pages with minimal backend overhead.
	\item Deterministic, seed-based question generation.
	\item Persistent storage of user attempts for future progression logic.
\end{itemize}

This architecture describes the \textbf{current implemented system}, not future optional changes.

\section{Fixed Technology Stack}

\subsection*{Application Runtime}

\begin{itemize}[leftmargin=*]
	\item \textbf{Framework}: Next.js 16 (App Router)
	\item \textbf{Language}: TypeScript
	\item \textbf{UI}: React 19 + Tailwind CSS 4
	\item \textbf{Server model}: Next.js route handlers + server components
\end{itemize}

\subsection*{Authentication}

\begin{itemize}[leftmargin=*]
	\item \textbf{Auth library}: better-auth
	\item \textbf{Adapter}: better-auth Prisma adapter
	\item \textbf{Mode}: Email + password with username plugin
\end{itemize}

\subsection*{Data Layer}

\begin{itemize}[leftmargin=*]
	\item \textbf{Database}: PostgreSQL
	\item \textbf{ORM}: Prisma (Prisma Client + PostgreSQL adapter)
	\item \textbf{Schema ownership}: \texttt{prisma/schema.prisma}
	\item \textbf{Seed strategy}: deterministic pattern upsert via \texttt{prisma/seed.ts}
\end{itemize}

\section{System Context}

At runtime, the system has four primary concerns:

\begin{enumerate}[leftmargin=*]
	\item \textbf{Web UI}: Landing, auth, dashboard, training, and overview pages.
	\item \textbf{Session \& Identity}: Sign-up/login/session validation via better-auth.
	\item \textbf{Training API}: Question retrieval and attempt persistence endpoints.
	\item \textbf{Persistence}: User, pattern, and attempt storage in PostgreSQL.
\end{enumerate}

\section{High-Level Module Architecture}

\subsection*{1) Presentation Layer (Next.js App Router)}

Main routes:

\begin{itemize}[leftmargin=*]
	\item \texttt{/} (landing page)
	\item \texttt{/login}, \texttt{/signup}
	\item \texttt{/dashboard}
	\item \texttt{/training}
	\item \texttt{/training/overview}
\end{itemize}

Characteristics:

\begin{itemize}[leftmargin=*]
	\item Dashboard and overview are server-rendered and gated by session checks.
	\item Training run UI is client-side to support real-time timer/input interaction.
\end{itemize}

\subsection*{2) API Layer (Route Handlers)}

\begin{itemize}[leftmargin=*]
	\item \texttt{/api/auth/[...all]} delegates to better-auth Next.js handler.
	\item \texttt{GET /api/training/calculations} returns generated questions from active level-1 patterns.
	\item \texttt{POST /api/training/attempts} validates and persists a single first-attempt event.
\end{itemize}

All training routes require an authenticated session.

\subsection*{3) Domain Logic Layer (Training Feature)}

Core modules under \texttt{src/features/training}:

\begin{itemize}[leftmargin=*]
	\item \texttt{question-generator.ts}: deterministic generator from \texttt{(domain, params, seed)}.
	\item \texttt{use-training-questions.ts}: fetch lifecycle for question batch loading.
	\item \texttt{use-training-run.ts}: run state machine (submit, wrong feedback, skip, advance, complete).
	\item \texttt{attempt-client.ts}: persistence call abstraction.
\end{itemize}

\subsection*{4) Infrastructure Layer}

\begin{itemize}[leftmargin=*]
	\item \texttt{src/lib/prisma.ts}: singleton Prisma client construction.
	\item \texttt{src/lib/auth.ts}: better-auth instance configuration.
	\item \texttt{src/lib/session.ts}: server-side session retrieval helper.
\end{itemize}

\section{Data Model Architecture}

\subsection*{Core Product Entities}

\begin{itemize}[leftmargin=*]
	\item \texttt{User}: identity/account owner.
	\item \texttt{Pattern}: reusable generator rule (domain, level, constraints, cutoff).
	\item \texttt{Attempt}: immutable first-attempt event for one question presentation.
	\item \texttt{UserDomainProgress}: per-user progress state by domain (prepared for progression logic).
\end{itemize}

\subsection*{Auth Support Entities}

\begin{itemize}[leftmargin=*]
	\item \texttt{Session}, \texttt{Account}, \texttt{Verification}
\end{itemize}

These are owned by better-auth and integrated through Prisma.

\subsection*{Key Enums}

\begin{itemize}[leftmargin=*]
	\item \texttt{Domain}: \texttt{ADD}, \texttt{SUB}, \texttt{MUL}, \texttt{DIV}
	\item \texttt{AttemptOutcome}: \texttt{CORRECT}, \texttt{WRONG}, \texttt{TIMEOUT}, \texttt{SKIPPED}
\end{itemize}

\subsection*{Important Relational Constraints}

\begin{itemize}[leftmargin=*]
	\item Unique pattern identity: \texttt{(domain, level, description)}.
	\item Unique user-progress tuple: \texttt{(userId, domain)}.
	\item Attempt indices optimized for user timeline, run replay, and pattern-level analysis.
\end{itemize}

\section{Training Runtime Flow}

\subsection*{A. Session Start}

\begin{enumerate}[leftmargin=*]
	\item User selects operations and count on dashboard.
	\item Browser navigates to \texttt{/training?operations=...\&count=...}.
	\item Client creates a \texttt{runId} (UUID) for this run.
\end{enumerate}

\subsection*{B. Question Batch Loading}

\begin{enumerate}[leftmargin=*]
	\item Client calls \texttt{GET /api/training/calculations}.
	\item Server validates session.
	\item Operations are mapped to domains; count is parsed with min/max bounds.
	\item Server fetches active patterns for selected domains and loads per-domain \texttt{UserDomainProgress.currentLevel} (defaulting to level 1 when absent).
	\item For each target question, server samples a playable level per domain using weighted progression distribution over \(\{1..L, L+1\}\), constrained by available seeded levels.
	\item Server samples a pattern from the selected domain and sampled level.
	\item For each question target, server samples a random seed and generates a deterministic question.
	\item Response contains \texttt{[\{patternId, seed, left, operator, right, answer\}]}.
\end{enumerate}

\subsection*{C. Per-Question Interaction}

For each question in the run:

\begin{itemize}[leftmargin=*]
	\item Timer starts at question load.
	\item First user submission is captured with response time.
	\item Correct first answer persists immediately and advances.
	\item Wrong first answer flashes feedback; user can retry locally.
	\item ``Don't know'' marks skip, reveals answer, then persists as skipped on continue.
\end{itemize}

\subsection*{D. Attempt Persistence \& Outcome Assignment}

On \texttt{POST /api/training/attempts}:

\begin{enumerate}[leftmargin=*]
	\item Request payload is structurally validated.
	\item Server loads pattern and regenerates expected question from \texttt{seed}.
	\item Server derives outcome:
	\begin{itemize}
		\item \texttt{SKIPPED} if skipped flag is true.
		\item \texttt{WRONG} if first submitted answer \(\neq\) expected answer.
		\item \texttt{CORRECT} if answer matches and \(firstResponseMs \le cutoffTimeMs\).
		\item \texttt{TIMEOUT} if answer matches but exceeds cutoff.
	\end{itemize}
	\item Immutable attempt row is written.
	\item Server evaluates recent attempts at the user's current domain level and applies promotion/demotion updates to \texttt{UserDomainProgress} when thresholds are met.
\end{enumerate}

\subsection*{E. Run Completion and Review}

\begin{enumerate}[leftmargin=*]
	\item On final question completion, client navigates to \texttt{/training/overview?runId=...}.
	\item Overview page loads attempts for the run in chronological order.
	\item UI renders summary counters and row-level results.
\end{enumerate}

\section{Deterministic Generation Design}

The architecture intentionally models a question by \texttt{(patternId, seed)}.

Benefits:

\begin{itemize}[leftmargin=*]
	\item No need to persist full question text at creation time.
	\item Reproducible verification of answers on the server.
	\item Lower storage overhead with auditable replay.
\end{itemize}

\subsection*{Generation Constraints by Domain}

\begin{itemize}[leftmargin=*]
	\item ADD: digit counts + exact carry count
	\item SUB: digit counts + exact borrow count (+ optional negatives)
	\item MUL: digit counts + multiplication carry count
	\item DIV: digit counts + exact integer constraints and quotient floor
\end{itemize}

Generator attempts are bounded to prevent unbounded loops for infeasible constraints.

\section{Auth \& Security Boundaries}

\begin{itemize}[leftmargin=*]
	\item Training APIs are session-protected.
	\item Attempt outcome is computed server-side; client cannot self-assign correctness.
	\item Persisted numeric values are converted into Prisma \texttt{Decimal} where relevant.
	\item Database write scope is minimized to required attempt fields.
\end{itemize}

\section{Operational Architecture}

\subsection*{Environment Configuration}

Minimum required runtime variables:

\begin{itemize}[leftmargin=*]
	\item \texttt{DATABASE\_URL}
\end{itemize}

\subsection*{Schema Lifecycle}

\begin{enumerate}[leftmargin=*]
	\item Apply Prisma migrations.
	\item Seed base patterns via \texttt{prisma/seed.ts}.
	\item Start application server.
\end{enumerate}

\subsection*{Client Generation}

Prisma client is generated into \texttt{src/generated/prisma}, then consumed by app code.

\section{Current State and Deliberate v1 Limits}

The architecture is production-shaped but intentionally narrow:

\begin{itemize}[leftmargin=*]
	\item Level selection uses weighted per-domain sampling based on \texttt{UserDomainProgress.currentLevel}.
	\item Dashboard progress widgets are currently mock visuals.
	\item Progression evaluation currently uses fixed v1 thresholds (promotion over last 10 solved; demotion over last 20 with unsolved rate threshold).
\end{itemize}

These are product-scope decisions, not architectural inconsistencies.

\section{Quality Attributes}

\subsection*{Strengths in v1}

\begin{itemize}[leftmargin=*]
	\item \textbf{Determinism}: server can always reconstruct expected answers.
	\item \textbf{Integrity}: outcomes derive from server logic, not client assertions.
	\item \textbf{Simplicity}: one monolithic codebase and one relational datastore.
	\item \textbf{Extensibility}: existing pattern/progress schema supports progression upgrades.
\end{itemize}

\subsection*{Known Tradeoffs}

\begin{itemize}[leftmargin=*]
	\item Tight coupling between web app and API (intentional in Next.js monolith).
	\item Minimal observability tooling in v1.
	\item Limited anti-abuse/rate-limiting controls on public endpoints.
\end{itemize}

\section{Reference Component Map}

\begin{longtable}{@{}p{0.28\linewidth}p{0.68\linewidth}@{}}
\toprule
Component & Responsibility \\
\midrule
\texttt{src/app/dashboard/page.tsx} & Session-gated configuration UI for operations and batch size. \\
\texttt{src/app/training/page.tsx} & Interactive run experience and run lifecycle orchestration. \\
\texttt{src/features/training/use-training-run.ts} & Run state machine for first-attempt capture and progression across questions. \\
\texttt{src/app/api/training/calculations/route.ts} & Auth-protected question batch generation from active patterns. \\
\texttt{src/app/api/training/attempts/route.ts} & Auth-protected attempt validation, outcome derivation, and persistence. \\
\texttt{src/features/training/question-generator.ts} & Deterministic constrained generator for ADD/SUB/MUL/DIV domains. \\
\texttt{src/lib/auth.ts} & better-auth initialization with Prisma adapter and plugins. \\
\texttt{src/lib/prisma.ts} & Shared Prisma client creation and non-production singleton reuse. \\
\texttt{prisma/schema.prisma} & Canonical data model, relations, enums, and indexing strategy. \\
\bottomrule
\end{longtable}

\section{Versioning Note}

This document is architecture baseline \textbf{v1} for the current repository state.
Future updates should revise this file when introducing:

\begin{itemize}[leftmargin=*]
	\item Dynamic per-user level selection in question retrieval.
	\item Automated progression writes into \texttt{UserDomainProgress}.
	\item Analytics/telemetry and operational hardening changes.
\end{itemize}

\end{document}
