\documentclass[11pt]{article}
\usepackage[a4paper,margin=1in]{geometry}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{amsmath}

\title{Mental Math Progression \& Level System (v1)}
\date{}

\begin{document}

\maketitle

\section{Vision}

\begin{itemize}[leftmargin=*]
    \item No global Elo system.
    \item Personalized progression per user.
    \item Each level corresponds to a \textbf{pattern family}.
    \item Progression based on:
    \begin{itemize}
        \item Relative speed (time-to-first-correct; median vs population)
        \item Skip behavior
    \end{itemize}
    \item No hard-coded time thresholds.
\end{itemize}

The system adapts automatically using population response-time percentiles.

\section{High-Level Structure}

Progression is separated by \textbf{Operation Domains}:

\begin{itemize}[leftmargin=*]
    \item Addition (ADD)
    \item Multiplication (MUL)
    \item Subtraction (SUB)
    \item Division (DIV)
\end{itemize}

Each domain has its own independent level ladder.

User progress is tracked per domain.

\section{Pattern-Based Levels}

Each level corresponds to a defined structural pattern.

A pattern is \textbf{not} a single question. It is a generator rule with structural constraints.

Each pattern contains:

\begin{itemize}[leftmargin=*]
    \item Operation
    \item Digit configuration
    \item Carry constraints (operation-specific)
    \item Other structural constraints (pattern-specific)
\end{itemize}

\subsection*{Operand Order Randomization}

For all patterns involving operands of different digit lengths, 
the order of the operands is randomized at generation time 
(e.g., $1$-digit $+$ $2$-digit or $2$-digit $+$ $1$-digit).
For equal-length operands (e.g., $1$-digit $+$ $1$-digit), 
order is naturally symmetric.

\subsection*{Example Patterns (Addition) --- first 5 levels}

\begin{itemize}[leftmargin=*]
    \item Level 1: 1-digit + 1-digit, no carry
    \item Level 2: 1-digit + 1-digit, with carry
    \item Level 3: 1-digit + 2-digit, no carry
    \item Level 4: 1-digit + 2-digit, with carry
    \item Level 5: 2-digit + 2-digit, no carry
\end{itemize}

\subsection*{Example Patterns (Multiplication) --- first 5 levels}

\begin{itemize}[leftmargin=*]
    \item Level 1: 1-digit $\times$ 1-digit, no carry
    \item Level 2: 1-digit $\times$ 1-digit, with carry
    \item Level 3: 1-digit $\times$ 2-digit, no carry
    \item Level 4: 1-digit $\times$ 2-digit, with carry
    \item Level 5: 1-digit $\times$ 2-digit, two carries
\end{itemize}

\subsection*{Example Patterns (Subtraction) --- first 5 levels}

\begin{itemize}[leftmargin=*]
    \item Level 1: 1-digit $-$ 1-digit, non-negative result (no borrow possible)
    \item Level 2: 2-digit $-$ 1-digit, no borrow
    \item Level 3: 2-digit $-$ 1-digit, with borrow (single borrow)
    \item Level 4: 2-digit $-$ 2-digit, no borrow
    \item Level 5: 2-digit $-$ 2-digit, with borrow (single borrow)
\end{itemize}


\subsection*{Example Patterns (Division) --- first 5 levels }

\begin{itemize}[leftmargin=*]
    \item Level 1: 1-digit $\div$ 1-digit, exact integer result
    \item Level 2: 2-digit $\div$ 1-digit, exact integer result
    \item Level 3: 2-digit $\div$ 2-digit, exact integer result
    \item Level 4: 3-digit $\div$ 1-digit, exact integer result
    \item Level 5: 3-digit $\div$ 2-digit, exact integer result
\end{itemize}



\section{Performance Model (v1: Hard-Coded Cutoff)}

Progression is based on:

\begin{itemize}[leftmargin=*]
    \item Whether the attempt is solved (correct answer submitted within $C_L$ milliseconds)
    \item If time-to-first-correct exceeds $C_L$, the attempt counts as unsolved regardless of correctness
\end{itemize}

For v1, each (domain, level) has a hard-coded \textbf{cutoff time} $C_L$.

This threshold is tuned manually and can later be replaced by population percentiles.

\section{Promotion Rule (Per Level)}

For a user currently on Level $L$ in a domain, evaluate the last $N$ completed attempts at Level $L$.

\begin{center}
$N = 10$
\end{center}

Promotion requires:

\begin{enumerate}[leftmargin=*]
    \item All of the last $N$ attempts are solved (correct and within $C_L$)
\end{enumerate}

If satisfied, unlock Level $L + 1$.

\section{Demotion Rule (Hysteresis)}

Evaluate the last $M$ attempts at the current level (including skips).

\begin{center}
$M = 20$
\end{center}

Demote if any of the following holds:

\begin{itemize}[leftmargin=*]
    \item Unsolved rate over the last $M$ attempts exceeds tolerance $S_{max}$
\end{itemize}

\begin{itemize}[leftmargin=*]
    \item $S_{max} = 0.25$ (25\%)
\end{itemize}

Only allow demotion after a minimum exposure at that level (20 attempts).

\section{Level Selection During Play}

When user is at Level $L$:

\begin{itemize}[leftmargin=*]
    \item 70\% questions from Level $L$
    \item 10\% from Level $L + 1$ (challenge)
    \item 20\% from Level $L - 1$ (confidence)
\end{itemize}

Within a level:

\begin{itemize}[leftmargin=*]
    \item Randomize seeds
    \item Randomize digit combinations
    \item Maintain structural constraints
\end{itemize}



\section{How Questions are stored}

\begin{enumerate}[leftmargin=*]
    \item Store \textbf{Pattern} rows (the generator rules).
    \item Generate question instances on demand using a RNG seed.
    \item Store each user interaction as an \textbf{Attempt} row.
\end{enumerate}

So, in practice, the ``question'' is represented by:

\begin{center}
\texttt{(patternId, seed)}
\end{center}

Optional later: add a \texttt{QuestionInstance} table only if you need explicit review queues or caching.

\subsection*{How a Pattern is stored in DB}

Each pattern family is one row in the \texttt{pattern} table.

Stable, indexed fields are stored as normal columns:

\begin{itemize}[leftmargin=*]
        \item \texttt{id}: unique pattern id
        \item \texttt{domain}: \texttt{ADD | MUL | SUB | DIV}
        \item \texttt{level}: ladder level within domain
        \item \texttt{description}: human-readable pattern label
        \item \texttt{cutoffTimeMs}: solved-time cutoff for v1
        \item \texttt{active}: enable/disable pattern
        \item \texttt{createdAt}, \texttt{updatedAt}: audit timestamps
\end{itemize}

Dynamic generation constraints are stored in:

\begin{itemize}[leftmargin=*]
        \item \texttt{params} (Prisma \texttt{Json}, PostgreSQL \texttt{JSONB})
\end{itemize}

Uniqueness rule in v1:

\begin{center}
\texttt{UNIQUE(domain, level, description)}
\end{center}

\subsection*{What \texttt{params} JSON looks like}

In v1, digit fields are stored as single integers (not lists),
because each pattern uses one fixed digit size per side.

\textbf{ADD example (Level 3: 1-digit + 2-digit, no carry)}

\begin{verbatim}
{
    "operation": "ADD",
    "leftDigits": 1,
    "rightDigits": 2,
    "carryCount": 0,
}
\end{verbatim}

\textbf{MUL example (Level 5: 1-digit x 2-digit, two carries)}

\begin{verbatim}
{
    "operation": "MUL",
    "leftDigits": 1,
    "rightDigits": 2,
    "carryCount": 2,
}
\end{verbatim}

\textbf{SUB example (Level 5: 2-digit - 2-digit, single borrow)}

\begin{verbatim}
{
    "operation": "SUB",
    "leftDigits": 2,
    "rightDigits": 2,
    "borrowCount": 1,
    "allowNegative": false
}
\end{verbatim}

\textbf{DIV example (Level 3: 2-digit / 2-digit, exact integer)}

\begin{verbatim}
{
    "operation": "DIV",
    "dividendDigits": 2,
    "divisorDigits": 2,
    "exactInteger": true,
    "minQuotient": 2
}
\end{verbatim}

The exact key set may vary by operation and version. The important rule is:
keep frequently queried fields in columns and operation-specific constraints in \texttt{params}.

\end{document}
